Towards Practical First-Order Model Counting

### 2025-07-09 doc/talk/talk.tex
Explain Crane

\documentclass{beamer}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{tikz}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{complexity}
\usepackage{fontawesome}
\usepackage{dashbox}
\usepackage[style=authoryear]{biblatex}

@@

% TODO: (maybe later, once I know more)
% 1) define FOMC: any number of domains, any number of predicates of any arity (do I need to define weights?)
% 2) introduce WFOMC algorithms
% 3) Introduce Greedy vs BFS?
\begin{frame}{First-Order Model Counting}
\end{frame}

\begin{frame}{The Problem with \textsc{Crane}}
  \begin{block}{A Solution Produced for the Bijection-Counting Problem}
  \begin{align*}
    f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\
    g(l, m) &= g(l-1, m) + mg(l-1, m-1)
  \end{align*}
  \end{block}
  \pause
  \begin{alertblock}{Issues}
    \begin{description}
      \item[Completeness:] what are the base cases of \structure{$g$}?
      \item[Usability:] how do I compute, e.g., \structure{$f(7, 7)$}?
    \end{description}
  \end{alertblock}
\end{frame}
% NOTE: after some simplifications
% NOTE: the first but not the best: it can also find a linear-time solution

>>>>

\documentclass{beamer}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{tikz}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{complexity}
\usepackage{fontawesome}
\usepackage{dashbox}
\usepackage[style=authoryear]{biblatex}
\usepackage{mathtools}

@@

\begin{frame}{More Formally: What Is the Input?}
  \[
    \forall x \in \alert<2>{\Gamma}\text{. }\forall y, z \in \alert<2>{\Delta}\text{. }P(x, y) \land P(x, z) \Rightarrow \alert<4>{y=z}
  \]
  \begin{block}{\mbox{\alert<2>{Many-Sorted} \alert<3>{Function-Free}
        First-Order Logic with \alert<4>{Equality}}}
  \begin{itemize}
    \item<5-> Any number of variables and constants
    % \item<5-> All variables are bound (by \alert{$\exists$} or
    %       \alert{$\forall$})
    \item<6-> \alert{$\exists$} and \alert{$\forall$} quantifiers can be nested
          arbitrarily deeply
    \item<6-> All domains are finite
          \begin{itemize}
            \item Solutions are functions that take domain sizes as input
          \end{itemize}
    % \item<6-> Predicates can have any arity
  \end{itemize}
  \end{block}
  \onslide<7->{
    \begin{block}{First-Order Model Counting (FOMC)}
      \begin{itemize}
        \item Each predicate acts like a \alert{subset}
        \begin{itemize}
          \item of a domain or
          \item of a Cartesian product of domains
        \end{itemize}
        \item Goal: count \alert{combinations of subsets} that satisfy the
              sentence
      \end{itemize}
    \end{block}
  }
\end{frame}

\begin{frame}{Previous Work:
    \textsc{Crane}~\color{gray}{\parencite{DBLP:conf/kr/DilkasB23}}}
  \begin{itemize}
    \item A knowledge compilation approach:
          \begin{itemize}
            \item \alert{Sentences} $\to$ labelled \alert{digraphs} $\to$
                  function-defining \alert{equations}
          \end{itemize}
    \item Capable of constructing recursive solutions
  \end{itemize}
  \pause
  \begin{exampleblock}{An Example Solution for Counting Bijections}
    \begin{align*}
      f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\
      g(l, m) &= g(l-1, m) + mg(l-1, m-1)
    \end{align*}
  \end{exampleblock}
  \pause
  \begin{alertblock}{Issues}
    \begin{description}
      \item[Completeness:] what are the base cases of \structure{$g$}?
      \item[Usability:] how do I compute, e.g., \structure{$f(7, 7)$}?
    \end{description}
  \end{alertblock}
\end{frame}
% NOTE: the first but not the best: it can also find a linear-time solution
% NOTE: this slide formulates the problem tackled by this work
